---
title: "Making a Wise Housing Choice"
author: "Michael Huang, Jiawei Wang, Huaqing Zhang, Qiushu Zhou"
date: "Oct 19, 2020"
output:
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, include=FALSE}
#install.packages("plyr")
#install.packages("aod")
#install.packages("lmtest")
library(tidyverse)
library(plyr)
library(dplyr)
library(aod)
library(lmtest)

gss <-read_csv("gss.csv")
attach(gss)
```
## House dezation

# Michael Huang, Jiawei Wang, Huaqing Zhang, Qiushu Zhou
# Oct 19, 2020

## Abstract

The study is based on the General Social Survey(GSS) data collected in 2017 and the goal of this study is to explore the relationship between the predictor variables age, total children, income of the respondent and the estimation variable, the household rented or owned by the respondent or a member of the household. We build a logistic regression model and the equation is $$logit(p) = log(p/(1-p))= -0.023 + 0.009*x_1 + 0.134*x_2 +0.186*x_3+0.870*x_4 + 1.185*x_5 + 1.336*x_6 + 1.797*x_7$$ (p is the probabilities of the house owned by the respondent). After applying a likelihood ratio test and diagnostic for the regression model, it shows that the model is well fitted. Our study shows that age, number of children and income of respondent are all valuable predictors for housing decision (rent or own) estimation.

## Introduction

  Human beings spend roughly 1/3 of their lifetime in bed and shelter is the most basic need of human beings. Making housing choices is very important since in most cases it is the biggest purchase in the lifetime.
  
  The goal of this study is to find out if age, number of children and income of respondent have an effect on the housing decision of owning or renting.  We will be using a logistic regression model since we are interested in whether the respondent owns their household or rents their household which is a binary response. To check how well our model fits., we used likelihood ratio test and diagnostic checks. After analyzing our model and graphs we have found that there is a relationship between our predictors and whether or not the respondent owns their housing.
  
## Data

  The data that we choose for our study is from the General Social Survey(GSS) in 2017, which is an annual survey tracking the changes within the Canadian society. Also, GSS data is used as important evidence for Canadian government to improve the well-being of Canadian society. A part of the data is directly collected from tax files and administrative files to reduce bias on some very personal information. Other information is collected from either electronic questionnaires or computer assisted telephone interviews. 
  
 The population of the data is all non-institutionalized persons 15 years of age or older, living in the 10 provinces of Canada. The survey uses a frame that combines landline and cellular telephone numbers from the Census and various administrative sources with Statistics Canada's dwelling frame. The sample of this survey is all respondents who completed this survey. Respondents are found from their registered address and telephone numbers. Non-response are not included in the sample. This is a sample survey with a cross-sectional design. This survey is very randomly distributed across Canada which means it creates less bias.  Moreover,  this survey  obtained better coverage of households with a telephone number.  There also exist 19 different variables, hence many different analyses can be done with this data. There are several drawbacks for this survey. Firstly, too many questions in this survey, which leads to many respondents not complete this survey. Also,  there are some quite personal questions that respondents did not answer them precisely, which result in the bias of the information. Moreover, the ‘NA’ value would exist in the collected data. 
 
In this report, we use 5 variables, caseid,  own rent, age, total children of  the respondent, and income of the respondent. Caseid, age and total children of the respondent are all numerical variables. In addition, own rent and income of the respondent are all categorical variables. 

 The dataset we use in this report is given below.
 

**Brief Data Table (Table 1)**
```{r,echo=FALSE}

gss_data <- select(gss, caseid, own_rent, age, total_children, income_respondent)

#Delete NA and "don't know" on own_rent and total_children
gss_data <- filter(gss_data, own_rent %in% c("Owned by you or a member of this household, even if it i...", "Rented, even if no cash rent is paid"))
cleaned_gss <- drop_na(gss_data, total_children) 
cleaned_gss
cleaned_gss$own_rent <- revalue(cleaned_gss$own_rent, c("Rented, even if no cash rent is paid"=0))
cleaned_gss$own_rent <- revalue(cleaned_gss$own_rent, c("Owned by you or a member of this household, even if it i..."=1))
cleaned_gss$income_respondent <- revalue(cleaned_gss$income_respondent, c("Less than $25,000"=1))
cleaned_gss$income_respondent <- revalue(cleaned_gss$income_respondent, c("$25,000 to $49,999"=2))
cleaned_gss$income_respondent <- revalue(cleaned_gss$income_respondent, c("$50,000 to $74,999"=3))
cleaned_gss$income_respondent <- revalue(cleaned_gss$income_respondent, c("$75,000 to $99,999"=4))
cleaned_gss$income_respondent <- revalue(cleaned_gss$income_respondent, c("$100,000 to $ 124,999"=5))
cleaned_gss$income_respondent <- revalue(cleaned_gss$income_respondent, c("$125,000 and more"=6))
cleaned_gss$own_rent <- as.numeric(cleaned_gss$own_rent)
cleaned_gss$income_respondent <- as.numeric(cleaned_gss$income_respondent)
```


## Model

The selected model for this study is the Logistic Regression Model which is a model that is used when the response variable has two possible outcomes like a yes or no question. In the survey there does not exist many numerical variables hence it is very hard to do simple linear regression. One of the intentions for this survey is to improve the well-being of Canadians, we choose housing as the response because having a nice place to live is one major factor of a person’s well-being. For this report, we want to explore the relationship between the variables age, total children, income and the outcome which is whether they rent or own a household. 

Before building the model, the data we used was modified. We are interested in whether the person rents or owns so every other option like "Don't know" and “NA” has been filtered out. Value for income has been changed into numbers from 1 to 6. Number 1 represents the lowest income range(Less than $25000) and so on. After cleaning this data, there are 20396 observations from age 15 to 80 in the dataset we use.

In order to fit a proper model and obtain relatively precise results, We made some re-arrangement on these data

As such we define the following binary response variable:

y=0 if the respondent **RENTS** the household.

y=1 if the respondent **OWNS** the household or a member of this household owned.


we use **logit(p) = log(p/(1-p))**, where p is the probabilities of y = 1 which refers to the household owned by respondent or a member of this household.


We will consider the following **SEVEN** potential predictor variables:

$x_1$ = age = The age of the respondent.

$x_2$ = total children = The number of children that the respondent has in total.

$x_3$ = The income of the respondent is from $25,000 to $49,999, noted as rank2.

$x_4$ = The income of the respondent is from $50,000 to $74,999, noted as rank3.

$x_5$ = The income of the respondent is from $75,000 to $99,999, noted as rank4.

$x_6$ = The income of the respondent is from $100,000 to $124,999, noted as rank5.

$x_7$ = The income of the respondent is from $125,000 and more, noted as rank6.

The reason why we choose these predictor variables is that different ages($x_1$) will lead to different attitudes towards household decisions,  the secure feeling a owned home can provide seems to become more important as age goes up.  Also the number of children ($x_2$) of the respondent will change the household decision of the respondent. If the respondent has more children, we predict that they are more likely to own a household. In addition, we predict that higher income of the respondent leads to higher possibilities for respondents to own a household.

We build a Logistic Regression Model:
$$logit(p)= \beta_0 + \beta_1x_1 + \beta_2x_2  + \beta_3x_3 + \beta_4x_4 + \beta_5x_5 + \beta_6x_6 + \beta_7x_7$$
In this model, we assume that once the income rank of the respondent is confirmed, then the other predictor about the income rank will automatically equal to 0. (e.g., if the income rank of the respondent is rank1, then $x_3$ = $x_4$ = $x_5$ = $x_6$ = $x_7$ = 0.)

We use glm() function to get the logistic regression model from Rstudio.

After we get the fitted model, we want to measure how well our model fits. We use a likelihood ratio test to measure whether the model with these 7 predictor variables fits significantly better than a model with just an intercept. In order to use the likelihood ratio test, we establish a null model which is a null logistic model with just an intercept. And we claim the null and alternative hypotheses.
  
  $H_0: \beta_1 = \beta_2 = \beta_3 = \beta_4 = \beta_5 = \beta_6 = \beta_7 = 0$

- The null model fits significantly better than the model with 7 predictor variables.
 
$H_A: \beta_1 \neq \beta_2 \neq \beta_3 \neq \beta_4 \neq \beta_5 \neq \beta_6 \neq \beta_7 \neq 0$

- The model with 7 predictor variables fits significantly better than the null model.

The test statistic for this hypothesis test is the difference in deviance for the two models, one is the model with 7 predictors and another is the null model with just an intercept. We use lrtest() function from package lmtest to do the likelihood ratio test and then we get the p-value.

To test the model we built, we also try to do a model diagnostic to check the assumptions of the logistic regression model.


## Results

From the logistic regression output from R (Figure 1) we get the fitted model is:
 $$logit(p) = -0.023 + 0.009*x_1 + 0.134*x_2 +0.186*x_3
                           +0.870*x_4 + 1.185*x_5 + 1.336*x_6 + 1.797*x_7$$

**Logistic Regression Output from R (Figure 1)**
```{r,echo=FALSE}
cleaned_gss$income_respondent <- factor(cleaned_gss$income_respondent)
logit_model <- glm(own_rent ~ age + total_children +
               as.factor(income_respondent), data=cleaned_gss, family="binomial")
summary(logit_model)
```                           
   
Likelihood ratio test shows that the test statistic is 1244.5, the chi-square of 1244.5 with 7 degrees of freedom and an associated p-value less than 0.001.

**Likelihood Ratio Test Output from R (Figure 2)**
```{r,echo=FALSE}
logit_null_model <- glm(own_rent ~ 1, data=cleaned_gss, family="binomial")
lrtest(logit_model, logit_null_model)
```                        

**Diagnostic for Logistic Regression Model Output from R (Figure 3)**

```{r,echo=FALSE}
par(mfrow = c(2,2))
plot(logit_model)
```

With the appearance of  the Residuals vs Fitted graph, it illustrates that there are not any distinctive patterns in the case. From the normal Q-Q graph, it shows that residuals are not lined very well on the straight dashed line, but it follows its trend as a whole. The Scale-Location graph shows if residuals are spread equally along the predictors. It is noted that one part of residuals are followed by the red smooth line with decreasing slope, and the residuals spread evenly. The Residuals vs Leverage graph helps us find if the influential cases exist, and we observe there are no any observations outside of Cook’s distance.

## Discussion

Our model studies the relationship between the outcome if respondents rent or own houses and some variables i.e. age, number of total children and income of respondents. We use the logistic regression model above to analyze the data. As the data that we selected in our study contains a part of categorical data, using the Logistic Regression can be a good choice and it’s predictor variables can be numerical or categorical. Since the categorical variable can use ‘dummy’ variable coding, we can perfectly analyze the data and make a conclusion about our thesis. 

According to the summary of the model, the fitted model is:

 $$logit(p) = -0.023 + 0.009*x_1 + 0.134*x_2 +0.186*x_3
                           +0.870*x_4 + 1.185*x_5 + 1.336*x_6 + 1.797*x_7$$
                           
From this fitted model we can conclude that the logistic regression coefficients give the change in the log odds of the outcome (logit(p)) for a one unit increase in the predictor variable. 

- When **age** of the respondent changes by 1 unit, the corresponding average change in logit(p) is **0.009**.

- When the **total children** of the respondent changes by 1 unit, the corresponding average change in logit(p) is **0.134**.

- The indicator variables for **income_respondent** shows that, if the income of respondent at **rank2** ($25,000 to $49,999), versus an income of rank1 (Less than $25,000), the average change in logit(p) is **0.186**. Similarly, if the income of respondent at **rank3** ($50,000 to $74,999), versus an income of rank1 (Less than $25,000), the average change in logit(p) is **0.870**, if the income of respondent at **rank4** ($75,000 to $99,999), versus an income of rank1 (Less than $25,000), the average change in logit(p) is **1.185**, if the income of respondent at **rank5** ($100,000 to $ 124,999), versus an income of rank1 (Less than $25,000), the average change in logit(p) is **1.336**, if the income of respondent at **rank6** ($125,000 and more), versus an income of rank1 (Less than $25,000), the average change in logit(p) is **1.797**

According to the likelihood ratio test, the chi-square of 1244.5 with 7 degrees of freedom and an associated p-value less than 0.001. It shows that we have strong evidence to reject H0 that the null model fits significantly better than the model with 7 predictor variables. Thus we can conclude that our logistic regression model(with 7 predictors) as a whole fits significantly better than the null model(with only intercept).

Based on our model diagnostics, since the Residuals vs Fitted and Normal Q-Q graph display,, one represents the linear relationships between the predictors and outcome variable (logit(p)), which met the linearity of the logistic regression model; the other shows normality assumption of the logistic regression model. Meanwhile, the Scale-Location and Residuals vs Leverage graph concluded that the data has equal constant and the model doesn’t have any cases with very high Cook’s distance. Hence, there are no influential cases affecting the established model. In summary, this model met the assumptions of the logistic regression model through the analysis of model diagnostics.

Based on the result of our analysis it is clear to conclude that age, number of children and income of the respondent are all important and valuable predictors for housing decision(rent or own) estimation. From our result we learned that making a housing decision actually relies on many factors. It is not a random decision that could be easily made or changed. There are some caveats, it should be very carefully used if the result is being applied to other countries. Also that there may be correlation between our chosen variable and another variable not included in this study which might affect the result. This model would be a good representation of developed countries. There wouldn’t be any major difference in income and number of children for all developed countries. However some developing countries might have a large number of children in the household and the income could also defer. For any future work it might be a good idea to compare data over the course of several years to see the change.
 

# Weaknesses

Since the study based on questionnaire as a method to collect the data, people can choose not to respond to the questions of survey. As can been seen from data 'gss.csv', there are many "NA" represented as "answer not available". Therefore, one of the weaknesses of our data is that it is not precise enough and there exists the missing information for the data. 
  
  Additionally, another needs to be noticed is that we used respondent's individual income as one of the factors. However, respondent's family income might be different from their personal income and could result in different effects on owned or rent house. We cannot provide thoughts on which of these two kinds of income is the most eligible "income factor" for our analysis on the outcome. Whereas in fact, except the interaction with family income, there are many other factors we can consider for our study objectives such as gender.
  
  In this project, we expect to discover the relationship between the outcome of people whether rent or owned house and the factors i.e. age, number of total children and their income. Meanwhile, we need to summarize if they will have the significant impacts on the outcome. If the missing value or incomplete information exist, we are not able to conclude on the significance of these selected variables to the outcome. Therefore, our final results might appear the bias of the conclusion. 

# Next Steps

To perfect our further study of the research, it is necessary to:

- Add some interactions
  Adding specific analysis on interactions between each variable can be considered in Next Steps. In our current study, we focus on how significant that each factor can influence the outcome 'own/rent'; however, we did not think about whether two factors in our study would have correlations with each other and finally we might get unexpected or mistaken results. 
  
- Add more variables such as sex, which can obtain differences between male and female.

- Do some follow-up survey:
  To verify our conclusion with the reality, we can set up another survey or questionnaire in order to ask participants "which factor do you think made you decide to own or rent a living place?". We can compare our analyzed results with respondent's autonomous answers to see whether our study on own/rent is rational or not.
  
  
## References

Achim Zeileis, Torsten Hothorn (2002). Diagnostic Checking in Regression Relationships. R News 2(3), 7-10. URL https://CRAN.R-project.org/doc/Rnews/

General Social Survey-Family(GSS). Statistics Canada.
https://www23.statcan.gc.ca/imdb/p2SV.pl?Function=getSurvey&SDDS=4501&lang=en&db=imdb&adm=8&dis=2

Hosmer, D. & Lemeshow, S. (2000). Applied Logistic Regression (Second Edition). New York: John Wiley & Sons, Inc.

Hadley Wickham, Romain François, Lionel Henry and Kirill Müller (2020). dplyr: AGrammar of Data Manipulation. R package version 1.0.2. https://CRAN.R-project.org/package=dplyr

Hadley Wickham (2011). The Split-Apply-Combine Strategy for Data Analysis. Journal of Statistical Software, 40(1), 1-29. URL http://www.jstatsoft.org/v40/i01/.

Lesnoff, M., Lancelot, R. (2012). aod: Analysis of Overdispersed Data. R package version 1.3.1, URL http://cran.r-project.org/package=aod

Logit Regression. UCLA: Statistical Consulting Group. https://stats.idre.ucla.edu/r/dae/logit-regression/

Long, J. Scott (1997). Regression Models for Categorical and Limited Dependent Variables. Thousand Oaks, CA: Sage Publications.

Wickham H, Averick M, Bryan J, Chang W, McGowan LD, François R, Grolemund G, Hayes A, Henry L, Hester J, Kuhn M, Pedersen TL, Miller E, Bache SM, Müller K, Ooms J, Robinson D, Seidel DP, Spinu V, Takahashi K, Vaughan D, Wilke C, Woo K, Yutani H (2019). “Welcome to the tidyverse.” Journal of Open Source Software, 4(43), 1686. doi: 10.21105/joss.01686.

RStudio Team (2020). RStudio: Integrated Development for R. RStudio, PBC, Boston, MA URL http://www.rstudio.com/.










